---
title: R extRemes Test with LOCA Downscaled CMIP-5 Model Output and some Horesplay
  with Ensembles (Example Using Precipitation)
output:
  html_notebook: default
  pdf_document: default
  html_document: default
---


Processing for Precipitation Daily Extremes Using LOCA Ensemble Output.

We will be usnig a Generalized Pareto Distribution to target the extreme values and that also requires a threshold to which the method is sensitive.  Temperature complicates this since we are looking at a goodly amount of non-stationarity as we move into the future.

A reference for the extRemes package can be found here

Gilleland, E., & Katz, R. (2016). extRemes 2.0: An Extreme Value Analysis Package in R. _Journal of Statistical Software_, __72__(8), 1-39. doi:http://dx.doi.org/10.18637/jss.v072.i08   ( PDF Avaialble at https://www.jstatsoft.org/article/view/v072i08 )


Warning Typos are Legion.



Load Required Libraries
```{r}

library(extRemes)  # extreme data analysis package

library(ncdf4)     # library for  processing netCDF data
library(lubridate) # library for processing dates and time
library(reshape2)  # library for manipulating data frames
library(beanplot)  # library for using bean plots for comparative plots
library(abind)     # library for merging multidimentional arrays
library(stringr)   # library for string manipulation

```



Let's start by identifying specific points, days and ensembles from which to pull data. 

It's nice to put this data up top to make changing locatons easier.

```{r}

location_name = "Rapid City"
location_abrv = 'KRAP'

target_lon  = -104. # degrees east
target_lat  =   44. # degrees north


# location_name = "Brookings"
# location_abrv = 'KBKX';

# target_lon  =  -96.8189167 # degrees east
# target_lat  =  44.3045278 # degrees north


```

And we should include our time periods.  Here, we have a baseline period and a future period

```{r}
period_span     = 30.
base_start      = 1976              # start year
base_end        = base_start + period_span-1 # end year  (just before our simulations diverge)

per1_start      = 2020             # start
per1_end        = per1_start + period_span-1 # end yvar2dear
```

And while I am up here I am defining a "wetting rain" even threshold as any rain event exceeding 0.1" per day

I tend to keep this value on reserve when working with rainfall whether I use it or not.  (In this case I am not...)

```{r}

wetting_rain_treshold = 0.1 * 25.4 # in -> mm (0.1 in = 2.54 )

```


The LOCA File for Rapid City can be downloaded from 

https://github.com/wjcapehart/SDMines_CEE_AES_Math_Ed_Resources/raw/master/Extreme_Event_Assessment/Data_For_Examples/KRAP_LOCA.nc


```{r}
# you will want to change the directory to point to where you are keeping the file.

URL_Root_Directory <- "~/GitHub/SDMines_CEE_AES_Math_Ed_Resources/Extreme_Event_Assessment/Data_For_Examples/"

# Here is the variable we want to extract
#  these should match the files names since the variable in the file and file name match

target_variable = "prec"
variable_name   = "Total Daily Precip"
variable_units  = "mm" # after post processing (starts as m/s)

threshold_value =  2.0 # This is needed for the Pareto Distribution.  

# get the extreme function to use

extreme_func = "GP"


# get URL files
LOCA_File  = paste(location_abrv,
                     "_LOCA.nc",
                     sep="")


# make the final URLs for extracting with the "paste" function

LOCA_URL = paste(URL_Root_Directory,  # string 1 to concatenate
                LOCA_File,         # string 2 to concatenate
                sep=""               # separation character ("" = none)
               )

```

We now open the files...

the output of these functions are "handles" by which we can reference the file
```{r}

nc.loca <- nc_open(filename = LOCA_URL)

```

To view the inventory (or "metadata") of a netCDF file you can "print()" the metadata

Let's do this with the RCP 8.5 file


```{r}

print(nc.loca)

```

Now for Time Control
    the RCP 8.5 runs go from 1920-2100 and have 2172 monthly time steps
    the RCP 4.5 runs go from 1920-2080 and have 1932 monthly time steps

Our data uses a 365-day year (no leap years) so we will doing this brute force.   

We are also only going to 2080.

```{r}

# get time values

time_hist<-ncvar_get(nc.loca, "time_hist")
tunits<-ncatt_get(nc.loca, "time_hist", attname="units")
tustr<-strsplit(tunits$value, " ")

time_reference_point_hist = unlist(tustr)[3]


time_futr <-ncvar_get(nc.loca, "time_futr")
tunits    <-ncatt_get(nc.loca, "time_futr", attname="units")
tustr     <-strsplit(tunits$value, " ")

# and tidy things up

time_reference_point_futr = unlist(tustr)[3]

```


Now for the easier coordinates to access 

```{r}
# create ensemble dimension

ensemble <- ncvar_get(nc.loca, "ensemble_model")

ensembles_to_use = ensemble[!str_detect(ensemble, "CNRM")]
ensembles_to_use = ensembles_to_use[!str_detect(ensembles_to_use, "GISS")]

n_ensembles = length(ensembles_to_use)

print(ensembles_to_use)

# import spatial coordinates

scenario <- ncvar_get(nc.loca, "RCP_scenario")
print(scenario)

scenario = c("RCP45","RCP85")


```

Pull Variable

```{r}

varid.hist = paste(target_variable,
                     "_hist",
                     sep="")
  
varid.futr = paste(target_variable,
                     "_futr",
                     sep="")
  
var2d.hist = ncvar_get(nc      = nc.loca,                      # netcdf file ID
                     varid   = varid.hist,            # variable name from file
                     verbose = FALSE                      # print diagnostic data
                     )                 # scaling temperature from K to DegC

var2d.futr = ncvar_get(nc    = nc.loca,                      # netcdf file ID
                     varid   = varid.futr,            # variable name from file
                     verbose = FALSE                      # print diagnostic data
                     )                  # scaling temperature from K to DegC


```

... Then we pull the RCP 8.5 scenario...

Assign dimension names to arrays.  these are strings including things that shoudl be numbers.

```{r}


dimnames(var2d.hist) <- list(ensemble,time_hist)

dimnames(var2d.futr) <- list(scenario,ensemble,time_futr)

size.hist2 = 2*20454*31
var2d.hist2  <- array(1:size.hist2, dim=c(2,31,20454))

var2d.hist2[1,,] = var2d.hist
var2d.hist2[2,,] = var2d.hist

dimnames(var2d.hist2) <- list(scenario,ensemble,time_hist)


var2d.hist = var2d.hist2
remove(var2d.hist2)

var3d = abind(var2d.hist, var2d.futr, along=3)


```




Now we flatten all of our fields into a single list or data frame by "flattening" the 2-D time/ensemble aray to a single list with the melt command

```{r}

variable = melt(data      = var3d,              # your array
              na.rm       = TRUE,                  # don't use missing values
              varnames    = c("Scenario","Ensemble","Time"),  # names of your two dimensions
              value.name  = "Variable")               # the final name of your aray value

variable$Time = as.Date(variable$Time, origin=time_reference_point_hist)
```
And we tidy things up once again

```{r}

remove(nc.loca, 
       varid.hist,
       varid.futr,
       URL_Root_Directory,
       LOCA_File,
       LOCA_URL,
       tunits,
       tustr,
       var3d,
       var2d.futr,
       var2d.hist)




```

We can also add some utility vectors to break things down by month, year and decade

```{r}
variable$month   = month(variable$Time)
variable$year   = year(variable$Time)
variable$day   = day(variable$Time)
variable$decade = trunc( variable$year/10. )*10



```


And we'll limit ourselves to the period between May and September

```{r}
#  variable = variable[(variable$month>=5 & variable$month<=9),]


```



It's not a bad idea to review the content of your data frame... 

```{r}
str(variable)

```

Now we're ready to do the analysis part.

we can break down the procedure to

1) par down the data vector by time period and ensemble number.
2) run and save our stats.

Here is a working sample template to get things started.

First we start extract our subset.

```{r}
ens_target = ensembles_to_use[1] # just use the first ensemble member
rpc_target = "RCP45"
subset = subset(variable,  
                (Scenario==rpc_target)  &
                (Ensemble==ens_target)  &
                ((year>=base_start) & 
                 (year<=base_end)   )    
                )

# always nice to see what your data looks like.

hist(x      = subset$Variable, # omitting zero temperatures days
     xlab   = paste(variable_name,                   # xaxis title
                    " (",
                    variable_units,
                    ")",
                    sep=""),
     main   = paste(location_name, 
                    paste("RCP8.5 Ensemble Member",ens_target),
                    sep = " "),
     freq   = TRUE,
     breaks = 100
     )



```

Now we run the fit command.

We will need a minimum threshold or the method fails.

Chose a reasonable value,  here we chose 2mm (0.08")

For reference, a wetting rain event is 0.1" or 2.54 mm

We are also using the Generalized Pareto method.

```{r}


fit_GP  = fevd(x          = Variable, 
               data       = subset, 
               verbose    = TRUE,
               threshold  = threshold_value,
               units      = variable_units,
               time.units = "365/year",
               type       = "GP",
               span       = period_span
               )



```

And we plot the results.  You would prefer for the lines to rest on the "1:1" through a reasonable period.  For a 30 year span, I'll consider 15 return yers to be acceptable.

```{r}
plot(fit_GP)
```

We can also extract specific return events.


```{r}

return_intervals = seq(from =             2,
                       to   = period_span/2,
                       by   =             1)


n_return_intervals = length(return_intervals) 

return_GP = return.level(x             = fit_GP,
                         return.period = return_intervals,
                         do.ci         = TRUE)

print(return_GP)

# tidy up the demo

remove(return_GP,
       fit_GP,
       subset)

```

To implement. Let's create the storage arrays  We will copy this over and over.

```{r}

# for each case/period create arrays for the return and CI bounds

# first for our base period

rcp45_base_period_return  = array(data= 0,
                                  dim = c(n_ensembles,
                                          n_return_intervals))

dimnames(rcp45_base_period_return) = list(ensembles_to_use, 
                                         return_intervals)

rcp45_base_period_ci_ub   = rcp45_base_period_return
rcp45_base_period_ci_lb   = rcp45_base_period_return

rcp85_base_period_return  = rcp45_base_period_return
rcp85_base_period_ci_ub   = rcp45_base_period_return
rcp85_base_period_ci_lb   = rcp45_base_period_return

# then for our test period(s)

rcp45_per1_period_return  = rcp45_base_period_return
rcp45_per1_period_ci_ub   = rcp45_base_period_return
rcp45_per1_period_ci_lb   = rcp45_base_period_return

rcp85_per1_period_return  = rcp45_base_period_return
rcp85_per1_period_ci_ub   = rcp45_base_period_return
rcp85_per1_period_ci_lb   = rcp45_base_period_return


```

create a loop that goes between the ensembles.

```{r}

ens_target_num = 0

for (ens_target in ensembles_to_use) {
  ens_target_num = ens_target_num + 1
  
  ######################################
  #
  # Base Period
  #
  
  # create segments for RCP 45 and RCP 85 for your period(s)
  
  
  ens_target
  
  rpc_target = "RCP45"

  subset = subset(variable,
                  (Ensemble==ens_target)  &
                  (Scenario==rpc_target)  &
                  ((year>=base_start) & 
                     (year<=base_end)   )
                       )
  
  # fit an extreme value dist to the data
  
  fit_GP_45 = fevd(x          = Variable,
                   data       = subset,
                   verbose    = FALSE,
                   threshold  = threshold_value,
                   units      = variable_units,
                   time.units = "365/year",
                   type       = "GP",
                   span       = period_span
                   )
  
   
  rpc_target = "RCP85"

  subset = subset(variable,
                  (Ensemble==ens_target)  &
                  (Scenario==rpc_target)  &
                  ((year>=base_start) & 
                     (year<=base_end)   )
                       )
  

  
  fit_GP_85 = fevd(x          = Variable,
                   data       = subset,
                    verbose    = FALSE,
                   threshold  = threshold_value,
                   units      = variable_units,
                   time.units = "365/year",
                   type       = "GP",
                   span       = period_span
                   )
  
  # calculate the return periods and load into the temporary storage arrays.
  
  return_GP = return.level(x             = fit_GP_45,
                           return.period = return_intervals,
                           do.ci         = TRUE)
  
  rcp45_base_period_ci_lb[ ens_target_num, ] = return_GP[ , 1]
  rcp45_base_period_return[ens_target_num, ] = return_GP[ , 2]
  rcp45_base_period_ci_ub[ ens_target_num, ] = return_GP[ , 3]

  
  return_GP = return.level(x             = fit_GP_85,
                           return.period = return_intervals,
                           do.ci         = TRUE)
   
  rcp85_base_period_ci_lb[ ens_target_num, ] = return_GP[ , 1]
  rcp85_base_period_return[ens_target_num, ] = return_GP[ , 2]
  rcp85_base_period_ci_ub[ ens_target_num, ] = return_GP[ , 3]
  
  # tidy up
  
  remove(subset,
         fit_GP_45,
         fit_GP_85,
         return_GP)

  #
  ######################################

  
  
  ######################################
  #
  #  Period #1
  #
  
  # create segments for RCP 45 and RCP 85 for your period(s)
  
 
  rpc_target = "RCP45"

  subset = subset(variable,
                  (Ensemble==ens_target)  &
                  (Scenario==rpc_target)  &
                  ((year>=per1_start) & 
                     (year<=per1_end)   )
                       )
  
  # fit an extreme value dist to the data
  
  fit_GP_45 = fevd(x          = Variable,
                   data       = subset,
                   verbose    = FALSE,
                   threshold  = threshold_value,
                   units      = variable_units,
                   time.units = "365/year",
                   type       = "GP",
                   span       = period_span
                   )
  
  rpc_target = "RCP85"

  subset = subset(variable,
                  (Ensemble==ens_target)  &
                  (Scenario==rpc_target)  &
                   ((year>=per1_start) & 
                     (year<=per1_end)   )
                       )
 
  fit_GP_85 = fevd(x          = Variable,
                   data       = subset,
                   verbose    = FALSE,
                   threshold  = threshold_value,
                   units      = variable_units,
                   time.units = "365/year",
                   type       = "GP",
                   span       = period_span
                   )
  
  # calculate the return periods and load into the temporary storage arrays.
  
  return_GP = return.level(x             = fit_GP_45,
                           return.period = return_intervals,
                           do.ci         = TRUE)
  
  rcp45_per1_period_ci_lb[ ens_target_num, ] = return_GP[ , 1]
  rcp45_per1_period_return[ens_target_num, ] = return_GP[ , 2]
  rcp45_per1_period_ci_ub[ ens_target_num, ] = return_GP[ , 3]

  
  return_GP = return.level(x             = fit_GP_85,
                           return.period = return_intervals,
                           do.ci         = TRUE)
   
  rcp85_per1_period_ci_lb[ ens_target_num, ] = return_GP[ , 1]
  rcp85_per1_period_return[ens_target_num, ] = return_GP[ , 2]
  rcp85_per1_period_ci_ub[ ens_target_num, ] = return_GP[ , 3]
  
  # tidy up
  
  remove(subset,
         fit_GP_45,
         fit_GP_85,
         return_GP)

  #
  ######################################

}

```

Convert output into neat and tidy data frames

```{r}


################################################
#
# let's start with the base period
#

# melt 2-d arrays into frames.

rcp45_base   = melt(data       = rcp45_base_period_return,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Return_Period")

rcp45_cib_lb = melt(data       = rcp45_base_period_ci_lb,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Lower_95_CI")

rcp45_cib_ub = melt(data       = rcp45_base_period_ci_ub,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Upper_95_CI")




rcp85_base   = melt(data       = rcp85_base_period_return,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Return_Period")

rcp85_cib_lb = melt(data       = rcp85_base_period_ci_lb,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Lower_95_CI")

rcp85_cib_ub = melt(data       = rcp85_base_period_ci_ub,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Upper_95_CI")



# combine the fields into a single data frame for each one.

rcp45_base$RCP45_Lower_95_CI = rcp45_cib_lb$RCP45_Lower_95_CI
rcp45_base$RCP45_Upper_95_CI = rcp45_cib_ub$RCP45_Upper_95_CI

rcp85_base$RCP85_Lower_95_CI = rcp85_cib_lb$RCP85_Lower_95_CI
rcp85_base$RCP85_Upper_95_CI = rcp85_cib_ub$RCP85_Upper_95_CI


# tidy up

remove(rcp85_base_period_ci_lb,
       rcp85_base_period_return,
       rcp85_base_period_ci_ub,
       rcp85_cib_lb,
       rcp85_cib_ub)

remove(rcp45_base_period_ci_lb,
       rcp45_base_period_return,
       rcp45_base_period_ci_ub,
       rcp45_cib_lb,
       rcp45_cib_ub)


#
################################################

################################################
#
# and we can continue with the test period
#


rcp45_per1   = melt(data       = rcp45_per1_period_return,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Return_Period")

rcp45_ci1_lb = melt(data       = rcp45_per1_period_ci_lb,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Lower_95_CI")

rcp45_ci1_ub = melt(data       = rcp45_per1_period_ci_ub,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP45_Upper_95_CI")




rcp85_per1   = melt(data       = rcp85_per1_period_return,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Return_Period")

rcp85_ci1_lb = melt(data       = rcp85_per1_period_ci_lb,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Lower_95_CI")

rcp85_ci1_ub = melt(data       = rcp85_per1_period_ci_ub,
                    varnames   = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "RCP85_Upper_95_CI")



# combine the fields into a single data frame for each one.

rcp45_per1$RCP45_Lower_95_CI = rcp45_ci1_lb$RCP45_Lower_95_CI
rcp45_per1$RCP45_Upper_95_CI = rcp45_ci1_ub$RCP45_Upper_95_CI

rcp85_per1$RCP85_Lower_95_CI = rcp85_ci1_lb$RCP85_Lower_95_CI
rcp85_per1$RCP85_Upper_95_CI = rcp85_ci1_ub$RCP85_Upper_95_CI


# tidy up

remove(rcp85_per1_period_ci_lb,
       rcp85_per1_period_return,
       rcp85_per1_period_ci_ub,
       rcp85_ci1_lb,
       rcp85_ci1_ub)

remove(rcp45_per1_period_ci_lb,
       rcp45_per1_period_return,
       rcp45_per1_period_ci_ub,
       rcp45_ci1_lb,
       rcp45_ci1_ub)


#
################################################

```


Plot return output

```{r}

# I am making my own colours

darkcyan = rgb(red   = 0.00,
               green = 0.50,
               blue  = 0.50, 
               alpha = 0.75)  # 1 = opaque; 0 = fully clear

cyan     = rgb(red   = 0.00,
               green = 1.00,
               blue  = 1.00, 
               alpha = 0.50)

darkblue = rgb(red   = 0.00,
               green = 0.00,
               blue  = 0.50, 
               alpha = 0.75)

blue     = rgb(red   = 0.00,
               green = 0.00,
               blue  = 1.00, 
               alpha = 0.50)


darkmag  = rgb(red   = 0.50,
               green = 0.00,
               blue  = 0.50, 
               alpha = 0.75)

magenta  = rgb(red   = 1.00,
               green = 0.00,
               blue  = 1.00, 
               alpha = 0.50)


darkred  = rgb(red   = 0.50,
               green = 0.00,
               blue  = 0.00, 
               alpha = 0.75)

red      = rgb(red   = 1.00,
               green = 0.00,
               blue  = 0.00, 
               alpha = 0.50)

beanplot(formula     = RCP85_Return_Period~return_intervals,  # formula selection for y axis
         data        = rcp85_base,                            # data frame to use
         col         = c(magenta,  # area fill                # Color Scheme
                         darkmag,  # lines
                         darkmag,  # outside of bean line
                         darkmag), # mean line
         border      = darkmag,                               # border color
         overallline = "median",                              # can't get rid of this dang thing
         beanlines   = "median",                              # use median for the "central value"
         main        = paste(location_name, 
                             " LOCA Return Intervals for ",         # title caption
                             base_start,
                            "-",
                            base_end,
                            sep=""),
         xlab        = "Return Period (Years)",               # xaxis title
         ylab        = paste("Maximum ",
                             variable_name,                   # yaxis title
                             " (",
                             variable_units,
                             ")",
                             sep=""),
         log         = "",                                    # overide log axis on y (it defaults to log for big datasets)
         xlim        = c(0,period_span/2),                    # x axis range
         ylim        = c(min(rcp45_base$RCP45_Lower_95_CI,    # y axis range
                             rcp85_base$RCP85_Lower_95_CI,
                             rcp45_per1$RCP45_Lower_95_CI,
                             rcp85_per1$RCP45_Lower_95_CI),
                         max(rcp45_base$RCP45_Upper_95_CI,    # y axis range
                             rcp85_base$RCP85_Upper_95_CI,
                             rcp45_per1$RCP45_Upper_95_CI,
                             rcp85_per1$RCP45_Upper_95_CI)
                        )
        )
        
        


beanplot(formula    = RCP45_Return_Period~return_intervals,
         data        = rcp45_base,
         add         = TRUE,
         beanlines   = "median",
         overallline = "median",
         col         = c(cyan,      # area fill
                         darkcyan,  # lines
                         darkcyan,  # outside of bean line
                         darkcyan), # mean line
         border      = darkcyan
        )    





```


Trying something new.  Now we can do this a "split" plot.

and for the future period.;; 

```{r}

# create a temporary array to hold just the return periods for simplicity

deleteme     = data.frame(Ensembles        = rcp45_base$Ensembles,
                          return_intervals = rcp45_base$return_intervals,
                          RCP45            = rcp45_per1$RCP45_Return_Period,
                          RCP85            = rcp85_per1$RCP85_Return_Period)

# melt the RCP45 and RCP85 columns into a new field. 
base_all     = melt(data       = deleteme,
                    varnames   = c("Ensembles","return_intervals","Scenario"),
                    id.vars    = c("Ensembles","return_intervals"),
                    na.rm      = TRUE,
                    value.name = "Return_Period_Values")

colnames(base_all) <- c("Ensembles","return_intervals","Scenario","Return_Period_Values")

# for this to work below, the program will need to search through ONE variable for the axis
# you can use two (or more) sub-categories.
#
# to do this you need to create a new variable in the data table that will have two
# parts, separated by a space character.
#
# this your x-axis + "series" category would be 
#
#  001 Cat1
#  002 Cat1
#  003 Cat1
#  001 Cat2
#  002 Cat3
# ... and so on.
#
# in this case we will have our return period year and then the RCP scenario. 
# (We could also do this for different periods for a single RCP scenario)
#
# thus...
#  02 RCP45 (we need that leading zero)
#  02 RCP85
#   ...
#  20 RCP45
#  20 RCP85


base_all$merged_period_scenarios = paste(sprintf("%2d", base_all$return_intervals),
                                         base_all$Scenario,
                                         sep = " ")

# tidy up!

remove(deleteme)

```
And now we can plot this one.  This will create a "split" bean plot which is about as complex and messy as I want to get right now.

```{r}



beanplot(formula     = Return_Period_Values~merged_period_scenarios,  # formula selection for y axis
         data        = base_all,                                     # data frame to use
         col         = list(c(blue, darkblue,darkblue,darkblue),  # same order as earlier for each series
                            c(red,  darkred,  darkred, darkred)), 
         border      = c(darkblue,darkred),                              
         overallline = "median",                              # can't get rid of this dang thing
         beanlines   = "median",                              # use median for the "central value"
         side        = "both",
         main        = paste(location_name,
                             " LOCA Return Intervals for ",         # title caption
                             per1_start,
                             "-",
                             per1_end,
                             sep=""),
         xlab        = "Return Period (Years)",               # xaxis title
         ylab        = paste("Maximum ",
                             variable_name,                   # yaxis title
                             " (",
                             variable_units,
                             ")",
                             sep=""),
         log         = "",                                    # overide log axis on y (it defaults to log for big datasets)
         xlim        = c(0,period_span/2),                    # x axis range
         ylim        = c(min(rcp45_base$RCP45_Lower_95_CI,    # y axis range
                             rcp85_base$RCP85_Lower_95_CI,
                             rcp45_per1$RCP45_Lower_95_CI,
                             rcp85_per1$RCP45_Lower_95_CI),
                         max(rcp45_base$RCP45_Upper_95_CI,    # y axis range
                             rcp85_base$RCP85_Upper_95_CI,
                             rcp45_per1$RCP45_Upper_95_CI,
                             rcp85_per1$RCP45_Upper_95_CI)
                        )
        )

legend("topleft", 
       fill   = c("blue", "red"), 
       legend = c("RCP 4.5", "RCP 8.5"))

```

